Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-01 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added appendix on tree-building algorithm (still needs to be revised)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2907 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 440b0d7f12318fbef4291da3d4e45f88d6e00225

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).
A more detailed description of the applied tree-building algorithm,
particularly of CTree, can be found in Appendix~\ref{app:tree}.

@@

%\newpage
%\section*{Appendix}
%\begin{appendix}
%\end{appendix}

@@

\begin{appendix}
\section{Tree-building algorithms}
\label{app:tree}
CTree, MOB
\end{appendix}
\end{document}

>>>>

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).
A more detailed description of the applied tree-building algorithm
can be found in Appendix~\ref{app:algorithm}.

@@

@@

\begin{appendix}

\section{Tree algorithm}
\label{app:algorithm}
This section provides a more detailed overview on the permutation-test-based CTree algorithm 
\citep{Hothorn+Hornik+Zeileis:2006}, 
specifically for circular data as applied to build the circular regression trees and forests 
in the presented case study. 
Alternatively, also the MOB algorithm, which is based on M-fluctuation tests, can be employed 
\citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.

\subsection*{CTree}
In the following the testing and splitting strategy is described for the root node of the tree
which contains the entire learning sample.
%consisting of circular observations denoted by $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
For a complete tree model, the same procedure is applied to all resulting child nodes with the 
corresponding subsamples.

First, employing the von Mises distribution, a distributional model $\text{vM}(Y; \mu,\kappa)$ 
is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
in Section~\ref{sec:prob_circ}.
In order to obtain a goodness-of-fit measurement of $\text{vM}(Y; \mu,\kappa)$ for each parameter 
and each observation the score function $s(\mu, \kappa, Y)$ is evaluated at the estimated location 
and concentration parameter $\hat{\mu}$ and $\hat{\kappa}$.
In a next step, a statistical test is employed to detect dependencies between the calculated score 
values
\begin{equation}
s(\hat{\mu}, \hat{\kappa}, y) = 
\begin{pmatrix} 
s(\hat{\mu}, \hat{\kappa}, y_1)_1 & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
\vdots & \vdots\\
s(\hat{\mu}, \hat{\kappa}, y_n)_1 & s(\hat{\mu}, \hat{\kappa}, y_n)_2
\end{pmatrix}
\end{equation}
and each possible split variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
In particular, a permutation test is applied to asses the following hypotheses:
\begin{align}
H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l.
\end{align}
%
The permutation test is based on the multivariate linear statistic
\begin{equation}
T_l = vec\left(\sum_{i=1}^n v_l(Z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, Y_i)\right)
\end{equation}
where $s(\hat{\mu}, \hat{\kappa}, Y_i) \in \mathbb{R}^{1\times 2}$. The form 
of the transformation function $v_l$ depends on the type of the split 
variable $Z_l$. If $Z_l$ is numeric then $v_l$ is simply the identity 
function $v_l(Z_{li}) = Z_{li}$ and therefore $T_l \in \mathbb{R}^2$ as 
the ``vec'' operator converts the $1 \times 2$ matrix into a $2$ column vector. 
If $Z_l$ is a categorical variable with $H$ categories then 
$v_l(Z_{li}) = (\I(Z_{li} = 1), \ldots, \I(Z_{li} = H))$ 
such that $v_l$ is a $H$-dimensional unit vector where the element corresponding to 
the value of $Z_{li}$ is $1$. In this case the statistic $T_l \in \mathbb{R}^{H \cdot 2}$
as the ``vec'' operator converts the $H \times 2$ matrix into a $H \cdot 2$ column vector 
by column-wise combination. Observations with missing values are excluded from the sums.

Using the conditional expectation $\mu_l$ and the covariance 
$\Sigma_l$ of $T_l$ as derived by \cite{Strasser+Weber:1999} 
the test statistic can be standardized. 
The observed multivariate linear statistic $t_l$ which is either a
$2$- or $2 \cdot H$-dimensional vector, depending on the scale of $Z_l$,
is mapped onto the real line by a univariate test statistic $c$.
In the application of this paper a quadratic form is chosen, such that
\begin{equation}
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
\end{equation}
where $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$. Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered ($c_{\text{max}}$).

\cite{Strasser+Weber:1999} showed that the asymptotic conditional 
distribution of the linear statistic $t_l$ is a multivariate normal with parameters
$\mu_l$ and $\Sigma_l$. Hence, the asymptotic conditional distribution of
$c(t_{l},\mu_{l},\Sigma_{l})$ 
is either normal
(for $c_{\text{max}}$) 
or 
$\chi^2$ (for $c_{\text{quad}}$).

The smaller the $p$-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.
%
After Bonferroni-adjusting the $p$-values it has to be assessed whether
any of the resulting $p$-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
is chosen as splitting variable. Otherwise no further split is made 
in this node as the stopping criterion of no $p$-values being below the 
significance level is fulfilled. This type of early-stopping in building
a tree is sometimes also referred to as ``pre-pruning''. For random forests
pre-pruning is often switched off by setting the significance level to $1$.

The breakpoint that leads to the highest discrepancy between score functions 
in the two resulting subgroups is selected as split point. 
This is measured by the linear statistic
\begin{equation}
T_{l^{\ast}}^{qr} = \sum_{i \in \mathcal{B}_{qr}} s(\hat{\mu}, \hat{\kappa}, Y_i)
\end{equation}
for $q \in \{1,2\}$ where $\mathcal{B}_{1r}$ and $\mathcal{B}_{2r}$ are the 
two new subgroups, without any particular ordering, that are defined by 
splitting in split point $r$ of variable $Z_{l^{\ast}}$.
The split point is then chosen as follows:
\begin{equation}
r^{\ast} = \argmin_{r} (\min_{q=1,2}(c(t_{l^{\ast}}^{qr},\mu_{l^{\ast}}^{qr},\Sigma_{l^{\ast}}^{qr})).
\end{equation}
One repeats the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal $p$-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.

%This permutation-test-based tree algorithm is presented in 
%\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
%A different framework to build a likelihood-based tree is provided by the MOB algorithm 
%which is based on M-fluctuation tests (\citealp{Zeileis+Hothorn+Hornik:2008}).

%\subsection{Teststatistic: CTree}\label{app:teststat_ctree}
%To measure the association of a response $Y$ and each possible split variable $Z_j$, $j=1,\ldots,J$,
%the CTree algorithm applies a linear test statistic $T_j$ which is excerpted from Section~3.1.
%(``Variable selection and stopping criteria'') of the original paper 
%\citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
%$$
%T_j(Y, Z_j, w) = \text{vec}\left( \sum_{i=1}^N w_i g_j(Z_{ji})h(Y_i)^{\top}\right)
%$$
%where $w$ are optional weights, $g_j: \mathcal{Z}_j \rightarrow \mathbb{R}^{P}$ is a nonrandom 
%transformation of the split variable $Z_j$ and the influence function 
%$h: \mathcal{Y} \times \mathcal{Y}^N \rightarrow \mathbb{R}^Q$ depends on the response 
%$Y = (Y_1, \ldots, Y_N)$ in a permutation symmetric way and is set to 
%$h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
%Moreover, by applying the ``vec'' operator the resulting
%$P~\times~Q$ matrix is converted into a $PQ$ column vector.
%Following \cite{Strasser+Weber:1999}
%the conditional expectation $\mu_j$ and covariance $\Sigma_j$ of a linear test statistic
%$T_j$ can be calculated and used to standardize an observed linear test statistic $t_j$ within a 
%function $c$ mapping into the real line.
%For example, the maximum of the absolute values of the standardized linear statistic
%$$
%c_{\max}(t_j,\mu_j,\Sigma_j) = \max \left| \frac{(t_j-\mu_j)}{\text{diag}(\Sigma_j)} \right|
%$$
%or a quadratic form
%$$
%c_{\text{quad}}(t_j, \mu_j, \Sigma_j) = (t_j - \mu_j)\Sigma_j^{+} (t_j - \mu_j)^{\top}
%$$
%can be considered where $\Sigma_j^{+}$ is the Moore-Penrose inverse of $\Sigma_j$.
%Since the asymptotic conditional distribution of a linear test statistic $T_j$ is a 
%multivariate normal with parameters $\mu_j$ and $\Sigma_j$ \citep{Strasser+Weber:1999}, 
%the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
%$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
%the corresponding $p$-values can be calculated easily.

\end{appendix}

\end{document}
